---
title: "GeneGraphs"
output:
  html_document: default
---
## Purpouse 

This script processes raw ribosome profiling data in order to visualize mapping of ribosome footprints to the transcriptome. The purpouse is to identify  alternative upstream translation initiation and quantify it in respect to canonical initiation.


```{r the set up}
#Loading libraries

library(rhdf5)
library(tidyr)
library(dplyr)
library(magrittr)
library(purrr)
library(ggplot2)

#Set up

nnt_gene<- 50

setwd("/Users/Ania/Desktop/Szkoła/4th year/Dissertation/gene_graphs/")

# prepare files, opens hdf5 file connection
dataset <- "G-Sc_2014"

hd_file <- "G-Sc_2014/output/WTnone/WTnone.h5"

hdf5file <- rhdf5::H5Fopen(hd_file) # filehandle for the h5 file

gene_names <- rhdf5::h5ls(hdf5file, recursive = 1)$name

# next step is to test with these 5 genes only instead of all genes
test_orfs<- c("YCR012W","YEL009C","YOR303W","YOL130W","YGR094W")

# 
orf_gff_file <- "G-Sc_2014/input/yeast_CDS_w_250utrs.gff3"

gff_df <- readGFFAsDf(orf_gff_file)

```

```{r set up}
head(gff_df)
```
## Functions

These functions are used to create a matrix of read lengths, positions and read counts. Get_Get5UTRstart, UTR5_length_table, table_modifications, UTR5_length_table, UTR5_length_table, UTR5_length set up parameters for matrix).

Function table_modifications doesn't work yet.

``` {r functions}

##Opening the gff file
readGFFAsDf <- purrr::compose(
  rtracklayer::readGFFAsGRanges,
  data.frame, 
  as_tibble,
  .dir = "forward" # functions called from left to right
)

##Finding the 5'UTR start position
#For GetGeneDataMatrixFunction (doesn't accept -ve values)
Get_Get5UTRstart <- function(name, gffdf, ftype="UTR5", fstrand="+") {
  gffdf %>% 
    dplyr::filter(type==ftype, Name == name, strand == fstrand) %>% 
    dplyr::pull(start)
}

#For TidyDataMatrix purpouses (renaming positions)
Tidy_Get5UTRstart <- function(name, gffdf, ftype="UTR5", fstrand="+") {
  gffdf %>% 
    dplyr::filter(type==ftype, Name == name, strand == fstrand) %>% 
    dplyr::pull(UTR5start)
}

#Creates new columns which will be useful for establishing boundaries of the matrix
table_modifications<- function(gffdf, ftype = "UTR5", fstrand = "+", End = "end", Start = "start") {
gffdf %>%
    filter(type==ftype, strand == fstrand) %>%
  mutate(UTR5start = gffdf$End *-1, UTR5L = seq(gffdf$Start:gffdf$End), UTR5_end = gffdf$Start-2 ) %>%
  select()
}

# Selects 5'UTR regions only for further modifications of the table
UTR5_length_table <- function(gffdf, ftype="UTR5", fstrand="+") {
  gffdf %>% 
    filter(type==ftype, strand == fstrand)} 

#Pulls 5'UTR length for GetGeneDataMatrix 
UTR5_length <- function(name, gffdf, ftype="UTR5", fstrand="+") {
  gffdf %>% 
    dplyr::filter(type==ftype, Name == name, strand == fstrand) %>% 
    dplyr::pull(UTR5L)
}

#Function for GeneDataMatrix
GetGeneDatamatrix <- function(gene, dataset, hdf5file) {
  hdf5file %>%
    rhdf5::H5Dopen(
      name = paste0("/", gene, "/", dataset, "/reads/data")
    ) %>%
    rhdf5::H5Dread() %>%
    return()
}

# Creates a matrix with number of columns that start at 5' UTR and finish at 50'th nt of the CDS
GetGeneDatamatrix5start <- function(gene, dataset, hdf5file, 
                                    Get_Get5UTRstart, UTR5full, nnt_gene) {
  data_mat_all <- GetGeneDatamatrix(gene, dataset, hdf5file)
  #I'm changing n_left5 to just 1st nt of 5'UTR
  n_left5 <- Get_Get5UTRstart # column to start from (5'end)
  n_right3 <- UTR5full + nnt_gene - 1 # column to end with (3'end) 
  data_mat_5start <- data_mat_all[, n_left5:n_right3]
  return(data_mat_5start)
}

```

## The GetGeneDatamatrix5start

This is what this function looks like when it is used.
At the moment, it works well for one gene but we might have to modify it for multiple genes.

Problems:
-what to do with the apply?
-should we modify it so it outputs one big tibble for all genes or individual tibble for each gene?

``` {r applying functions}

example <-lapply(test_orfs[1],
                function(gene)
                  GetGeneDatamatrix5start(gene,
                          dataset,
                          hdf5file,
                          Get_Get5UTRstart(gene, gff_df),
                          UTR5full = UTR5_length(gene, gff_df),
                          nnt_gene)
)  #change gffdf to x

head(example)
```
## Tidying data

The main struggle here is to get rid of the reduce() function, as we want to achieve the opposite; instead of summarizing data into metagenomic data, we want to get access into each gene separately.

I do not really know how to make it run for each gene, but here are some of the steps to tidy the data

``` {r tidying data}
YCR012W<-as_tibble(interesting$YCR012W, .name_repair ="minimal")

UTR5_YCR012W<-Tidy_Get5UTRstart(test_orfs[1], gff_df2)

positions <- UTR5_YCR012W : (UTR5_YCR012W +ncol(wtf)-1)

try<-YCR012W %>% 
  set_colnames(positions) %>%
  gather(key="Position", value = "Counts") %>%
  group_by(Position) %>%
  summarise(Counts=sum(Counts)) %>%
  arrange(try$Position)
 #For some reason, it doesn't actually arrange it

try


```

## Plotting 

Plotting with ggplot. Unfortunately the data isn't arranged (somehow that function doesn't always work).

```{r plot bit}

try_plot<-ggplot(
  data=try,) +
  geom_histogram(aes(x=as.numeric(Position), y=Counts), stat="identity") + 
  theme_light() +
  labs(y= "Read count", x = "Position (mRNA)") +
  scale_x_discrete(limit = c(-100,-50,0,50))

try_plot

ggsave("try_plot_example", device = "png")

```